Program Transformations for miniKanren

### 2020-07-17 papers/miniKanren/text/abstract.tex
Camera-ready miniKanren'20

\begin{abstract}
  We explore partial deduction for \mk{}: a specialization technique aimed at improving the performance of a relation in the given direction.
  We describe a novel approach to specialization of \mk{} based on partial deduction and supercompilation.
  On several examples, we demonstrate issues which arise during partial deduction.
\end{abstract}

>>>>

\begin{abstract}
  We explore partial deduction, an advanced specialization technique aimed at improving the performance of a relation in the given direction, in the context of \mk{}.
  On several examples, we demonstrate issues which arise during partial deduction of relational programs.
  We describe a novel approach to specialization of \mk{} based on partial deduction and supercompilation.
  Although the proposed approach does not give the best results in all cases, we view it as a stepping stone towards the efficient optimization of \mk{}.

\end{abstract}

### 2020-07-17 papers/miniKanren/text/conclusion.tex
spell- and grammarly- checked
Camera-ready miniKanren'20

We compared some of the specialization techniques for \mk{}.

There seem to be no one good technique.

>>>>

In this paper, we discussed some issues which arise in partial deduction of a relational programming language, \mk{}.
We presented a novel approach to partial deduction which uses less-branching heuristic to select the most suitable relation call to unfold at each step of driving.
We compared this approach to the earlier implementation of conjunctive partial deduction and the implementation of CPD equipped with the new less-branching heuristic.

The conservative partial deduction improved the execution time of all relations while the implementations of CPD degraded the performance of some of them.
However, CPD equipped with the less-branching heuristic improved the execution time of one relation the most compared with the other specializers.
We conclude that there is still no one good technique which definitely speeds up every relational program.
More research is needed to develop models to predict performance of relations, these models can further be used in specialization.

### 2020-07-17 papers/miniKanren/text/evaluation.tex
Camera-ready miniKanren'20

In our study we compared the CPD adaptation for \mk{} and the new non-conjunctive partial deduction.
We have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check wether it can improve the quality of the specialization.

We used the following 4 programs to test the specializers on.
\begin{itemize}
  \item Two implementations of an evaluator of logic formulas.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

@@

The relation \lstinline{eval$^o$} describes evaluation of a subset of first-order logic formulas in a given substitution.
It has 3 arguments.
The first argument is a list of boolean values which serves as a substitution.
The $i$-th value of the substitution is the value of the $i$-th variable.
The second argument is a formula with the following abstract syntax.
A formula is either a \emph{variable} represented with a Peano number, a \emph{negation} of a formula, a \emph{conjunction} of two formulas or a \emph{disjunction} of two formulas.
The third argument is the value of the formula in the given substitution.

All examples of \mk{} relations in this paper are written in \oc{}\footnote{\oc{}: statically typed \mk{} embedding in \ocaml{}. The repository of the project: \url{https://github.com/JetBrains-Research/OCanren}} syntax.
We specialize the \lstinline{eval$^o$} relation to synthesize formulas which evaluate to \lstinline{^true}\footnote{An arrow lifts ordinary values to the logic domain}.
To do so, we run the specializer for the goal with the last argument fixed to \lstinline{^true}, while the first two arguments remain free variables.
Depending on the way the \lstinline{eval$^o$} is implemented, different specializers generate significantly different residual programs.

@@

One possible implementation of the evaluator in the syntax of \oc{} is presented in listing~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean operations.

@@

Note, that the calls to boolean relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} are placed last within each conjunction.
This poses a challenge to the CPD-based specializers.
Conjunctive partial deduction unfolds relation calls from left to right, so when specializing this relation for running backwards (i.e. considering the goal \lstinline{eval$^o$ subst fm ^true}), it fails to propagate the direction data onto recursive calls of \lstinline{eval$^o$}.
Knowing that \lstinline{res} is \lstinline{^true}, we can conclude that in the call \lstinline{and$^o$ v w res} variables \lstinline{v} and \lstinline{w} have to be \lstinline{^true} as well.
There are three possible options for these variables in the call \lstinline{or$^o$ v w res} and one for the call \lstinline{not$^o$}.
These variables are used in recursive calls of \lstinline{eval$^o$} and thus restrict the result of driving them.
CPD fails to recognize this, and thus unfolds recursive calls of \lstinline{eval$^o$} applied to fresh variables.
It leads to over-unfolding, big residual programs and less than optimal performance.

The non-conjunctive partial deduction first unfolds those calls which are selected with the heuristic.
Since exploring boolean operations makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way non-conjunctive partial deduction treats this program is the same as it treats the other implementation in which boolean operations are moved to the left, as shown in listing~\ref{eval:fst}.
This program is easier for CPD to transform which demonstrates how unequal is the behavior of CPD for the similar programs.

@@

Depending on the way a relation is implemented, it may take different number of driving steps to reach the point when any useful information is derived through its unfolding.
Partial deduction tries to unfold every relation call unless it is unsafe, but not all relation calls serve to restrict the search space and thus should be unfolded.
In the implementation of \lstinline{eval$^o$} boolean operations can effectively restrict variables within the conjunctions and should be unfolded until they do.
But depending on the way they are implemented, different number of driving steps should be performed for that.
The simplest way to implement these relations is with a table as demonstrated with the implementation of \lstinline{not$^o$} in listing~\ref{not:table}.
It is enough to unfold such relation calls once to derive useful information about variables.

@@

The other way to implement boolean operations is via one basic boolean relation such as \lstinline{nand$^o$} which is in turn has a table-based implementation (see listing~\ref{not:nando}).
It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Non-conjunctive partial deduction drives the selected call until it derives useful substitutions for the variables involved while CPD with deterministic unfolding may fail to do so.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
    \begin{lstlisting}[label={not:nando}, caption={Implementation of boolean operation via \lstinline{nand}}, captionpos=b, frame=tb]
  let not$^o$ x y = nand$^o$ x x y

@@

In our study we considered two implementations of \lstinline{eval$^o$}, one we call \lstinline{plain} and the other --- \lstinline{last}, and compared how specializers behave on them.
The \lstinline{plain} relation uses table-based boolean operations and places them further to the left in each conjunction.
The relation \lstinline{last} employs boolean operations implemented via \lstinline{nand$^o$} and place them at the end of each conjunction.
These two programs are complete opposites from the standpoint of CPD.

We measured time necessary to generate $10000$ formulas over two variables which evaluate to \lstinline{^true}.
We compared the results of specialization of the goal \lstinline{eval$^o$ subst fm ^true} by our implementation of CPD, the new non-conjunctive partial deduction, and the CPD modified with the branching heuristic.
Our evaluation confirmed that CPD behaves very differently on these two implementations of the same relation: the running time of the specialized \lstinline{last} relation is almost 6 times as high as the running time of the specialized \lstinline{plain} relation.
The running time of two programs generated with our novel non-conjunctive partial deduction is very close and it is a little bit better than the best by CPD.
CPD with the branching heuristic still constructs residual programs of different quality.
The results are shown in table~\ref{tbl:eval}.

Besides evaluator of logic formulas we also run the transformers on the relation \lstinline{unify} which searches for a unifier of two terms and a relation \lstinline{isPath} specialized to search for paths in the graph.
These two relations are described in paper~\cite{lozov2019relational} so we will not go into too much details here.

The \lstinline{unify} relation was executed to find a unifier of two terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk{} program fails to terminate on this goal in 300 seconds.
On this example the most performant is the program generated by CPD (2.35 seconds) while the program generated by adding branching heuristic also fails to terminate in 300 seconds.
The non-conjunctive partial deduction shows some improvement with the residual program terminated within 15 seconds.
While driving this program, non-conjunctive partial deduction does too much unfolding which negatively impact the running time as compared to CPD.

The last test executed \lstinline{isPath} relation to search for 5 paths in the graph with 20 vertices and 30 edges.
On this program non-conjunctive partial deduction showed better transformation results than CPD, although the difference is not that drastic.

All evaluation results are presented in the table~\ref{tbl:eval}.
Each column correspond to the relation being run as described above.
The row marked ``Original'' contains the running time of the original \mk{} relation before specialization, ``CPD'' and ``Non-CPD'' correspond to conjunctive and non-conjunctive partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c||c||c}
                   & last & plain & unify & isPath \\
  \hline\hline
  Original         & >60.00s & >60.00s & >300.00s & 19.86s \\
  \hline
  CPD              & 31.31s & 5.46s & 2.35s & 4.66s \\
  \hline
  Non-CPD          & 4.99s  & 5.05s & 14.90s & 3.00s \\
  \hline
  Branching        & 17.21s  & 6.17s & >300.00s & N/A \\
  \hline
  \end{tabular}

@@

@@

% The unification of two terms $t$ and $u$ is searching for a substitution $\theta$ such that $t \theta = u \theta$, $\theta$ is called a unifier.
% We search for any unifier, not necessarily most specific.
% The details on this benchmark can be found in the earlier paper CITE.
% This example demonstrates how too much unfolding can be introduced with the non-conjunctive partial deduction.

>>>>

We implemented the new conservative partial deduction\footnote{The repository of the \mk{} specialization project: \url{https://github.com/kajigor/uKanren_transformations}. Access date: 17.07.2020.} and compared it with the CPD adaptation for \mk{} of~\cite{lozov2019relational}.
We have also employed the branching heuristic instead of the deterministic unfolding in the CPD to check whether it can improve the quality of the specialization.

We used the following programs to test the specializers.
\begin{itemize}
  \item Two implementations of an evaluator of logic formulas.
  \item A program to compute a unifier of two terms.
  \item A program to search for paths of a specific length in a graph.
\end{itemize}

@@

The relation \lstinline{eval$^o$} describes an evaluation of a subset of first-order logic formulas in a given substitution.
It has 3 arguments.
The first argument is a list of boolean values which serves as a substitution.
The $i$-th value of the substitution is the value of the $i$-th variable.
The second argument is a formula with the following abstract syntax.
A formula is either a \emph{variable} represented with a Peano number, a \emph{negation} of a formula, a \emph{conjunction} of two formulas or a \emph{disjunction} of two formulas.
The third argument is the value of the formula in the given substitution.

All examples of \mk{} relations in this paper are written in \oc{}\footnote{\oc{}: statically typed \mk{} embedding in \ocaml{}. The repository of the project: \url{https://github.com/JetBrains-Research/OCanren}. Access date: 17.07.2020.} syntax.
We specialize the \lstinline{eval$^o$} relation to synthesize formulas which evaluate to \lstinline{^true}\footnote{An arrow lifts ordinary values to the logic domain.}.
To do so, we run the specializer for the goal with the last argument fixed to \lstinline{^true}, while the first two arguments remain free variables.
Depending on the way the \lstinline{eval$^o$} is implemented, different specializers generate significantly different residual programs.

@@

One possible implementation of the evaluator in the syntax of \oc{} is presented in Listing~\ref{eval:last}.
Here the relation \lstinline{elem$^o$ subst v res} unifies \lstinline{res} with the value of the variable \lstinline{v} in the list \lstinline{subst}.
The relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} encode corresponding boolean operations.

@@

Note, that the calls to boolean relations \lstinline{and$^o$}, \lstinline{or$^o$}, and \lstinline{not$^o$} are placed last within each conjunction.
This poses a challenge to the CPD-based specializers.
Conjunctive partial deduction unfolds relation calls from left to right, so when specializing this relation for running backwards (i.e. considering the goal \lstinline{eval$^o$ subst fm ^true}), it fails to propagate the direction data onto recursive calls of \lstinline{eval$^o$}.
Knowing that \lstinline{res} is \lstinline{^true}, we can conclude that in the call \lstinline{and$^o$ v w res} variables \lstinline{v} and \lstinline{w} have to be \lstinline{^true} as well.
There are three possible options for these variables in the call \lstinline{or$^o$ v w res} and one for the call \lstinline{not$^o$}.
These variables are used in recursive calls of \lstinline{eval$^o$} and thus restrict the result of driving them.
CPD fails to recognize this, and thus unfolds recursive calls of \lstinline{eval$^o$} applied to fresh variables.
It leads to over-unfolding, big residual programs and poor performance.

The conservative partial deduction first unfolds those calls which are selected with the heuristic.
Since exploring boolean operations makes more sense, they are unfolded before recursive calls of \lstinline{eval$^o$}.
The way conservative partial deduction treats this program is the same as it treats the other implementation in which boolean operations are moved to the left, as shown in Listing~\ref{eval:fst}.
This program is easier for CPD to transform which demonstrates how unequal the behaviour of CPD for similar programs is.

@@

Depending on the way a relation is implemented, it may take a different number of driving steps to reach the point when any useful information is derived through its unfolding.
Partial deduction tries to unfold every relation call unless it is unsafe, but not all relation calls serve to restrict the search space and thus not every relation call should be unfolded.
In the implementation of \lstinline{eval$^o$} boolean operations can effectively restrict variables within the conjunctions and should be unfolded until they do.
But depending on the way boolean operations are implemented, different number of driving steps should be performed for that.
The simplest way to implement these relations is with a table as demonstrated with the implementation of \lstinline{not$^o$} in Listing~\ref{not:table}.
It is enough to unfold such relation calls once to derive useful information about variables.

@@

The other way to implement boolean operations is via one basic boolean relation such as \lstinline{nand$^o$} which has, in turn, a table-based implementation (see Listing~\ref{not:nando}).
It will take several sequential unfoldings to derive that variables \lstinline{v} and \lstinline{w} should be \lstinline{^true} when considering a call \lstinline{and$^o$ v w ^true} implemented via a basic relation.
Conservative partial deduction drives the selected call until it derives useful substitutions for the variables involved.
CPD with deterministic unfolding may fail to derive useful substitutions.

\begin{figure*}[!h]
  \centering
  \begin{minipage}{0.7\textwidth}
    \begin{lstlisting}[label={not:nando}, caption={Implementation of boolean operations via \lstinline{nand}}, captionpos=b, frame=tb]
  let not$^o$ x y = nand$^o$ x x y

@@

In our study, we considered two implementations of \lstinline{eval$^o$}, one we call \lstinline{plain} and the other --- \lstinline{last}, and compared how specializers behave on them.
The \lstinline{plain} relation uses table-based boolean operations and places them further to the left in each conjunction.
The relation \lstinline{last} employs boolean operations implemented via \lstinline{nand$^o$} and places them at the end of each conjunction.
These two programs are complete opposites from the standpoint of CPD.

We measured the time necessary to generate $1000$ formulas over two variables which evaluate to \lstinline{^true}.
We compared the results of specialization of the goal \lstinline{eval$^o$ subst fm ^true} by our implementation of CPD, the new conservative partial deduction, and the CPD modified with the less-branching heuristic.
Our evaluation confirmed that CPD behaves very differently on these two implementations of the same relation.
CPD improves the execution time of the \lstinline{plain} relation, however CPD performs too much unfolding of the  \lstinline{last} relation which is why the specialized relation \lstinline{last} fails to terminate in under 10 seconds.
The execution time of two programs generated with the novel conservative partial deduction is very similar and it is a little bit better than the best by CPD.
CPD with the less-branching heuristic constructs residual programs of different quality, worsening the execution time for both implementations.
The results are shown in table~\ref{tbl:eval}.

Besides the evaluator of logic formulas we also run the transformers on the relation \lstinline{unify}, which searches for a unifier of two terms, and the relation \lstinline{isPath} specialized to search for paths in a graph.
These two relations are described in paper~\cite{lozov2019relational} so we will not go into too many details here.

The \lstinline{unify} relation was executed to find a unifier of the terms $f(X, X, g(Z, t))$ and $f(g(p, L), Y, Y)$.
The original \mk{} program fails to terminate on this goal in 30 seconds.
On this example, the most performant is the program generated by conservative partial deduction (0.96 seconds).

The last test executed the \lstinline{isPath} relation to search for 5 paths in a graph with 20 vertices and 30 edges.
The original \mk{} program fails to terminate on this goal in 30 seconds.
On this program, CPD with branching heuristic showed much better transformation result than both CPD and conservative partial deduction, although all specialized versions show improvement as compared with the original relation.

All evaluation results are presented in the table~\ref{tbl:eval}.
Each column corresponds to the relation being run as described above.
The row marked ``Original'' contains the execution time of the original \mk{} relation before specialization, ``CPD'' and ``ConsPD'' correspond to conjunctive and conservative partial deduction respectively while ``Branching'' is for the CPD modified with the branching heuristic.

\begin{table}
  \centering
  \begin{tabular}{c||c|c||c||c}
                   & last  & plain & unify  & isPath \\ \hline\hline
  Original         & 1.06s & 1.84s & ---    & ---    \\ \hline
  CPD              & ---   & 1.13s & 14.12s & 3.62s  \\ \hline
  ConsPD           & 0.93s & 0.99s & 0.96s  & 2.51s  \\ \hline
  Branching        & 3.11s & 7.53s & 3.53s  & 0.54s  \\ \hline
  \end{tabular}

@@

% \begin{table}
%   \centering
%   \begin{tabular}{c||c|c||c||c}
%                    & last   & plain  & unify    & isPath  \\ \hline\hline
%   Original         & 1.29s & 1.61s & >30.00s & 25.84s \\ \hline
%   CPD              & 1.77s & 0.17s & 0.34s   & 3.54s  \\ \hline
%   Non-CPD          & 1.07s & 0.98s & 0.41s   & 1.68s  \\ \hline
%   Branching        & 0.51s & 0.18s & >30.00s & N/A     \\ \hline
%   \end{tabular}

%   \caption{Evaluation results}
%   \label{tbl:eval}
% \end{table}

% \begin{table}
%   \centering
%   \begin{tabular}{c||c|c||c||c}
%                    & last & plain & unify & isPath \\
%   \hline\hline
%   Original         & >60.00s & >60.00s & >300.00s & 19.86s \\
%   \hline
%   CPD              & 31.31s & 5.46s & 2.35s & 4.66s \\
%   \hline
%   Non-CPD          & 4.99s  & 5.05s & 14.90s & 3.00s \\
%   \hline
%   Branching        & 17.21s  & 6.17s & >300.00s & N/A \\
%   \hline
%   \end{tabular}

%   \caption{Evaluation results}
%   \label{tbl:eval}
% \end{table}

@@

% The unification of two terms $t$ and $u$ is searching for a substitution $\theta$ such that $t \theta = u \theta$, $\theta$ is called a unifier.
% We search for any unifier, not necessarily most specific.
% The details on this benchmark can be found in the earlier paper CITE.
% This example demonstrates how too much unfolding can be introduced with the conservative partial deduction.

### 2020-07-17 papers/miniKanren/text/intro.tex
Related works
Camera-ready miniKanren'20

The core feature of the family of relational programming languages \mk{}\footnote{\mk{} language web site: \url{http://minikanren.org}} is their ability to run a program in different directions.
Having specified a relation for adding two numbers, one can also compute the substraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running \emph{backwards} a relational interpreter for some language.
In general, it is possible to create a solver for a recognizer by translating it into \mk{} and running in the appropriate direction~\cite{lozov2019relational}.

@@

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} showed that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk{} programs.
Unfortunately, it may also not affect the running time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro{} have been studied before~\cite{leuschel2002logic}.
The ideas described there are aimed at left-to-right evaluation strategy of \pro{}.
Since the search in \mk{} is complete, it is safe to reorder come relation calls within the goal for better performance.
While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper we show on examples some issues which conjunctive partial deduction faces.
We also describe a novel approach to partial deduction of a relational programming language \mk{}.
We compare it to the existing specialization algorithms on several programs and discuss why some \mk{} programs run slower after specialization.

>>>>

The core feature of the family of relational programming languages \mk{}\footnote{\mk{} language web site: \url{http://minikanren.org}. Access date: 17.07.2020.} is their ability to run a program in different directions.
Having specified a relation for adding two numbers, one can also compute the subtraction of two numbers or find all pairs of numbers which can be summed up to get the given one.
Program synthesis can be done by running \emph{backwards} a relational interpreter for some language.
In general, it is possible to create a solver for a recognizer by translating it into \mk{} and running in the appropriate direction~\cite{lozov2019relational}.

@@

Specialization or partial evaluation~\cite{jonesbook} is a technique aimed at improving the performance of a program given some information about it beforehand.
It may either be a known value of some argument, its structure (i.e. the length of an input list) or, in case of a relational program, --- the direction in which it is intended to be run.
An earlier paper~\cite{lozov2019relational} showed that \emph{conjunctive partial deduction}~\cite{de1999conjunctive} can sometimes improve the performance of \mk{} programs.
Unfortunately, it may also not affect the running time of a program or even make it slower.

Control issues in partial deduction of logic programming language \pro{} have been studied before~\cite{leuschel2002logic}.
The ideas described there are aimed at left-to-right evaluation strategy of \pro{}.
Since the search in \mk{} is complete, it is safe to reorder some relation calls within the goal ahead-of-time for better performance.
While sometimes conjunctive partial deduction gives great performance boost, sometimes it does not behave as well as it could have.

In this paper, we show on examples some issues which conjunctive partial deduction faces.
We also describe \emph{conservative partial deduction} --- a novel specialization approach for the relational programming language \mk{}.
We compare it to the existing specialization algorithms on several programs and discuss why some \mk{} programs run slower after specialization.

### 2020-07-17 papers/miniKanren/text/related.tex
Camera-ready miniKanren'20

Specialization is an attractive technique aimed to improve the performance of a program if some of its arguments are known statically.
It is studied for functional, imperative and logic programing and comes in different forms: partial evaluation~\cite{jonesbook} and partial deduction~\cite{lloyd1991partial}, supercompilation~\cite{soerensen1996positive}, distillation~\cite{hamilton2007distillation} and many more.

The heart of supercompilation-based techniques is \emph{driving}~---~a symbolic execution of a program through all possible execution paths.
The result of driving is so-called \emph{process tree} where nodes correspond to \emph{configurations} which present computation state, for example, a term in case of pure functional programming languages.
Each path in the tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variable values, equalities, and disequalities and precomputing of all deterministic semantic evaluation steps, i.e. combining of consecutive process tree nodes with no branching, also known as \emph{deforestation}~\cite{TODO}.
When the tree is constructed, the resulting, or \emph{residual}, program can be extracted from the process tree by the process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure termination in supercompilation.
If the whistle signalled during the construction of some branch, then something should be done to ensure termination.
The most common approaches include either stopping driving the infinite branch completely (no specialization is done in this case and the source code is blindly copied into the residual program) or folding the process tree to fold in a \emph{process graph}.
The main instrument to perform such a folding is \emph{generalization}, i.e. abstracting away some computed data about the current term which makes folding possible.
One source of infinite branches is consecutive recursive calls to the same function with an accumulating parameter: by unfolding such a call further one can only increase the term size which leads to nontermination.
The accumulating parameter can be removed by replacing the call with its generalization.
There are several ways to ensure process correctness and termination, the most common being \emph{homeomorphic embedding}~\cite{TODO} used as a whistle and most-specific generalization of terms.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a function with accumulating parameter, and the first is not an instance of the second, then one may construct a new, generalized, node such that both of original nodes are instances of the last one, then one of the initial nodes is replaced with the generalized one.

While supercompilation generally improves the behavior of input programs and distillation can even provide superlinear speedup, there are no ways to predict the effect of specialization on a given program in general case.
What is worse, they rarely consider the residual program efficiency from the point of view of the target language evaluator.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at transformation-time.
Other criteria like the size of the generated program or possible optimizations and execution cost of different language constructions by the target language evaluator are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the real speedup of any given program on concrete examples even disregarding problems above because of the complexity of the transformation algorithm.
The worst case for partial evaluation is when all static variables are used in a dynamic context, and there is some advice on how to implement a partial evaluator as well as a target program so that specialization really improved its performance~\cite{jonesbook,bulyonkov84}.
There is lack of research in determining the classes of programs which transformers would definitely speed up.

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for left-to-right evaluation strategy of \pro{}.
CPD constructs a tree which models goal evaluation and is similar to SLD-NF tree, then a residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the local control determines the shape of the residual programs, while the global control ensures that every relation which can be called in the residual program is indeed defined.
The leaves of local control trees become nodes of the global control tree.
CPD analyses these nodes at the global level and runs local control for all those which are new.

At the local level CPD examines a conjunction of atoms by considering each atom one-by-one from left to right.
An atom is \emph{unfolded} if it is deemed safe, i.e. a whistle based on homeomorphic does not signal for the atom.
When an atom is unfolded, a clause whose head can be unified with the atom is found, and a new node is added into the tree where the atom in the conjunction is replaced with the body of that clause.
If there are more than one suitable head, then several branches are added into the tree which correspond to the disjunction in the residualized program.
An adaptation of CPD for the \mk{} programming language is described in~\cite{lozov2019relational}.

The most well-behaved strategy of local control in CPD for \pro{} is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if only one suitable clause head exist for it with the one exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a non-deterministic atom is the leftmost within a conjunction, it is most likely to be unfoldled and introduce many new relation calls within the conjunction.
We believe this is the core problem with CPD which limits its power when applied to \mk{}.
The strategy of unfolding atoms from left to right is reasonable in the context of \pro{} because it mimics the way programs in \pro{} execute.
But it often leads to larger global control trees and, as a result, bigger, less efficient programs.
The evaluation result of a \mk{} program does not depend on the order of atoms (relation calls) within a conjunction, thus we believe a better result can be achieved by selecting a relation call which can restrict the number of branches in the tree.
We describe our approach which implements this idea in the next section.

>>>>

Specialization is an attractive technique aimed to improve the~performance of a~ program if some of its arguments are known statically.
Specialization is studied for functional, imperative and logic programing and comes in different forms: partial evaluation~\cite{jonesbook} and partial deduction~\cite{lloyd1991partial}, supercompilation~\cite{soerensen1996positive}, distillation~\cite{hamilton2007distillation}, and many more.

The heart of supercompilation-based techniques is \emph{driving}~---~a~ symbolic execution of a~ program through all possible execution paths.
The result of driving is a~ \emph{process tree} where nodes correspond to \emph{configurations} which represent computation state.
For example, in the~case of pure functional programming languages, the~computational state might be a~ term.
Each path in the~tree corresponds to some concrete program execution.
The two main sources for supercompilation optimizations are aggressive information propagation about variables' values, equalities and disequalities, and precomputing of all deterministic semantic evaluation steps.
The latter process, also known as \emph{deforestation}, means combining of consecutive process tree nodes with no branching.
When the~tree is constructed, the~resulting, or \emph{residual}, program can be extracted from the~process tree by the~process called \emph{residualization}.
Of course, process tree can contain infinite branches.
\emph{Whistles} --- heuristics to identify possibly infinite branches --- are used to ensure supercompilation termination.
If a~ whistle signals during the~construction of some branch, then something should be done to ensure termination.
The most common approaches are either to stop driving the~infinite branch completely (no specialization is done in this case and the~source code is blindly copied into the~residual program) or to fold the~process tree to a~ \emph{process graph}.
The main instrument to perform such a~ folding is \emph{generalization}.
Generalization, abstracting away some computed data about the~current term, makes folding possible.
One source of infinite branches is consecutive recursive calls to the~same function with an accumulating parameter: by unfolding such a~ call further one can only increase the~term size which leads to nontermination.
The accumulating parameter can be removed by replacing the~call with its generalization.
There are several ways to ensure process correctness and termination,
most-specific generalization (anti-unification) and \emph{homeomorphic embedding} ~\cite{Higman52,Kruskal60} as a~ whistle being the~most common.
% When two dangerously similar nodes are encountered,
% For example, two consecutive recursive calls of a~ function with accumulating parameter, and the~first is not an instance of the~second, then one may construct a~ new, generalized, node such that both of original nodes are instances of the~last one, then one of the~initial nodes is replaced with the~generalized one.

While supercompilation generally improves the~behaviour of input programs and distillation can even provide superlinear speedup, there are no ways to predict the~effect of specialization on a~ given program in the~general.
What is worse, the~efficiency of residual program from the~target language evaluator point of view is rarely considered in the~literature.
The main optimization source is computing in advance all possible intermediate and statically-known semantics steps at program transformation-time.
Other criteria, like the~size of the~generated program or possible optimizations and execution cost of different language constructions by the~target language evaluator, are usually out of consideration~\cite{jonesbook}.
It is known that supercompilation may adversely affect GHC optimizations yielding standalone compilation more powerful~\cite{SCBE,TCES} and cause code explosion~\cite{SCHC}.
Moreover, it may be hard to predict the~real speedup of any given program on concrete examples even disregarding the~problems above because of the~complexity of the~transformation algorithm.
The worst-case for partial evaluation is when all static variables are used in a~ dynamic context, and there is some advice on how to implement a~ partial evaluator as well as a~ target program so that specialization indeed improves its performance~\cite{jonesbook,bulyonkov84}.
There is a~ lack of research in determining the~classes of programs which transformers would definitely speed up.

Conjunctive partial deduction~\cite{de1999conjunctive} makes an effort to provide reasonable control for the~left-to-right evaluation strategy of \pro{}.
CPD constructs a~ tree which models goal evaluation and is similar to a~ SLDNF tree, then a~ residual program is generated from this tree.
Partial deduction itself resembles driving in supercompilation~\cite{gluck1994partial}.
The specialization is done in two levels of control: the~local control determines the~shape of the~residual programs, while the~global control ensures that every relation which can be called in the~residual program is indeed defined.
The leaves of local control trees become nodes of the~global control tree.
CPD analyses these nodes at the~global level and runs local control for all those which are new.

At the~local level, CPD examines a~ conjunction of atoms by considering each atom one-by-one from left to right.
An atom is \emph{unfolded} if it is deemed safe, i.e. a~ whistle based on homeomorphic embedding does not signal for the~atom.
When an atom is unfolded, a~ clause whose head can be unified with the~atom is found, and a~ new node is added into the~tree where the~atom in the~conjunction is replaced with the~body of that clause.
If there is more than one suitable head, then several branches are added into the~tree which corresponds to the~disjunction in the~residualized program.
An adaptation of CPD for the~\mk{} programming language is described in~\cite{lozov2019relational}.

The most well-behaved strategy of local control in CPD for \pro{} is \emph{deterministic unfolding}~\cite{leuschel1997advanced}.
An atom is unfolded only if precisely one suitable clause head exists for it with the~single exception: it is allowed to unfold an atom non-deterministically once for one local control tree.
This means that if a~ non-deterministic atom is the~leftmost within a~ conjunction, it is most likely to be unfolded and to introduce many new relation calls within the~conjunction.
We believe this is the~core problem with CPD which limits its power when applied to \mk{}.
The strategy of unfolding atoms from left to right is reasonable in the~context of \pro{} because it mimics the~way programs in \pro{} execute.
But it often leads to larger global control trees and, as a~ result, bigger, less efficient programs.
The evaluation result of a~ \mk{} program does not depend on the~order of atoms (relation calls) within a~ conjunction, thus we believe a~ better result can be achieved by selecting a~ relation call which can restrict the~number of branches in the~tree.
We describe our approach which implements this idea in the~next section.
